---
title: "Cheatsheet for the minimum stats/plotting needed for UG Psychology."
date: ~
output: tufte::tufte_handout
---


# Pre-reqs

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, include=T, message=F, warning = F)
library(tidyverse)
library(pander)
library(broom)
```

  
# Working with data

Have a look at the data:

```{r}
airquality %>% glimpse
```


```{r}
airquality %>% 
  head
```



## Selecting columns, filtering rows and sorting

Select rows and or columns; sorting; show first 3:

```{r}
gapminder::gapminder %>% 
  filter(lifeExp > 80) %>% 
  arrange(year) %>% 
  select(year, country) %>% 
  head(3)
```



## Group by, and summarise

```{r}
lme4::sleepstudy %>% 
  group_by(Subject) %>% 
  summarise(mean(Reaction)) %>% 
  head(5)
```



## Reshaping long to wide

Filter, select and spread wide:

```{r}
gapminder::gapminder %>% 
  filter(year>2000) %>% 
  select(year, country, lifeExp) %>% 
  spread(year, lifeExp) %>% 
  head(5)
```



## Reshaping wide to long

Gather from wide to long form:

```{r}
iris %>%
  gather(variable, value, -Species) %>% 
  head
```


Or using `data.table` and `melt`:

```{r}
library(data.table)
iris %>%
  melt(id.var="Species") %>% 
  sample_n(10)
```




## Joining two separate datasets:

```{r}
demographics <- data_frame(person = c(1,2,3), age = c(23, 25, 21))
responses <- data_frame(person=c(1,1,1, 2,2,2), trial = c(1,2,3, 1,2,3), rt = c(230,245,232, 343,356,374))

left_join(responses, demographics, by="person")
```





## Creating combinations for conditions for an experiment design

```{r}
design <- expand.grid(condition = c("A", "B"), 
            stimulus = LETTERS[1:3], 
            participant = 1:20)

design %>% head(10)
```

Randomising presentation within-participant:

```{r}
design %>% 
  mutate(i = runif(n())) %>% 
  arrange(participant, i) %>% 
  mutate(trial = row_number()) %>% 
  select(participant, trial, everything(), -i) %>% 
  head(10)
```




\clearpage

# Plotting


```{r}
gapminder::gapminder %>% 
  ggplot(aes(year, lifeExp, group=year)) + 
  geom_boxplot()
```


```{r}
gapminder::gapminder %>% 
  ggplot(aes(year, lifeExp, color=continent)) + 
  geom_point(position="jitter", alpha=.5, size=.75) + 
  geom_smooth() + 
  ylab("Life expectancy")
```





```{r}
lme4::sleepstudy %>% 
  ggplot(aes(Days, Reaction)) + 
  geom_point(alpha=.3, size=.75) +
  geom_smooth(se=F, method="lm") + 
  facet_wrap(~Subject)
```



```{r}
mtcars %>% 
  mutate(am = factor(am, labels=c("Manual", "Automatic")), cyl=factor(cyl)) %>% 
  ggplot(aes(am, mpg, color=cyl, group=cyl)) +
  stat_summary(geom="pointrange", fun.data=mean_cl_boot) + 
  stat_summary(geom="line", fun.data=mean_cl_boot) + 
  xlab("")
```





\clearpage


# Correlation


As a plot:

```{r, message=F, warning=F}
GGally::ggpairs(airquality)
```



A table in APA format:

```{r}
apaTables::apa.cor.table(airquality)
```

As a `data_frame`:

```{r}
air.cor <- psych::corr.test(airquality)
air.cor$r %>% 
  broom::tidy()
```





\clearpage

# Chi Sq

First make a table of counts with `xtabs`:

```{r}
geartable <- xtabs(~gear+cyl, data=mtcars)
geartable
```


Then run the test on the table:

```{r}
chisq.test(geartable)
```




\clearpage

# Linear models

Examples of simple regression; multiple regression with interaction; multiple regression with interaction of continuous by categorical predictors. 

```{r}
m1 <- lm(mpg ~ wt , data=mtcars)
m2 <- lm(mpg ~ wt*cyl , data=mtcars)
m3 <- lm(mpg ~ wt*factor(cyl) , data=mtcars)
```



## Coefficient summaries from models

```{r}
summary(m1)
```


Extracting model coefficients as a dataframe:

```{r}
m2 %>% 
  broom::tidy(conf.int=T) %>% 
  pander
```


Extracting model fit details too:

```{r}
m2 %>% 
  broom::glance() %>% 
  pander
```




## Anova tables


```{r}
car::Anova(m3, type=3)
```


Format in APA format:

```{r}
apaTables::apa.aov.table(m3)
```



## Post-hoc tests

Run a suitable model:

```{r}
m4 <- lm(conformity ~ fcategory*partner.status, data=car::Moore)
car::Anova(m4)
```


And plot the marginal means:

```{r}
lsmeans::lsmeans(m4, ~partner.status*fcategory) %>% 
  broom::tidy() %>% 
  ggplot(aes(partner.status, estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_pointrange() +
  facet_wrap(~paste0("fcategory=", fcategory)) 
```



And extract the pairwise test statistics with FDR correction:

```{r}
pairs(lsmeans::lsmeans(m4, ~partner.status*fcategory), adjust="fdr") %>% 
  broom::tidy() %>% 
  pander(caption="Pairwise tests, with FDR correction.")
```




## Repeated measures Anova/Mixed Models

These models are (roughly) equivalent ways of doing the similar things:

```{r}
data(obk.long, package = "afex")

rm1 <- afex::aov_4(value ~ treatment * gender + (phase * hour | id), 
        data = obk.long, observed = "gender")

rm2 <- lmerTest::lmer(value ~ treatment * gender * hour * phase + (1 | id), data=obk.long)
```


The Anova tables for each model:

```{r}
anova(rm1)

lmerTest::anova(rm2)
```


Extract means/contrasts for groups:

```{r}
lsmeans::lsmeans(rm1, ~treatment)
pairs(lsmeans::lsmeans(rm1, ~treatment))

lmerTest::difflsmeans(rm2, "treatment")
```


Plotting the results of the model:

```{r}
lsmeans::lsmeans(rm1, ~treatment*gender) %>% 
  broom::tidy() %>% 
  ggplot(aes(treatment, estimate, 
             ymin=conf.low, ymax=conf.high, 
             color=gender, group=gender)) + 
  geom_pointrange(position=position_dodge(width = .1))
```



Within a mixed-model framework, we can also allow for varying effects of phase/hour by participant:

```{r}
rm3 <- lmerTest::lmer(value ~ treatment * gender * hour * phase + (phase + hour | id), data=obk.long)
lmerTest::anova(rm3)
```


And we can test whether the random effects are different to zero:

```{r}
lmerTest::rand(rm3)
```





